
---
title: "Comparing the Utility of Traits in Explaining NPP Variation"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

This analysis compares the utility of no traits, qualitative traits, and quantitative traits in explaining the variation in net primary productivity (NPP) in a simulated community over 20 years.

## Define Species and Traits

First, we define the species and their traits, including both quantitative and qualitative traits.

```{r define-traits}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(lme4)
library(stats)
library(forecast)
library(lmtest)

# Set seed for reproducibility
set.seed(42)

# Define species names
species <- paste0("Species_", 1:10)

# Define quantitative traits and assign random values
quantitative_traits <- data.frame(
  Species = species,
  Growth_Rate = runif(10, 0.1, 1.0),
  Photosynthetic_Efficiency = runif(10, 0.1, 1.0),
  Leaf_Area_Index = runif(10, 1, 10),
  Water_Use_Efficiency = runif(10, 0.1, 1.0)
)

# Define qualitative traits and assign random values
qualitative_traits <- data.frame(
  Species = species,
  Plant_Type = sample(c("Algae", "Vascular"), 10, replace = TRUE),
  Shade_Tolerance = sample(c("High", "Medium", "Low"), 10, replace = TRUE)
)

# Combine traits into one data frame
traits <- quantitative_traits %>%
  left_join(qualitative_traits, by = "Species")

# Define carrying capacity for each species
traits$Carrying_Capacity <- runif(10, 5000, 10000)

# Print traits
print(traits)
```

## Set Initial Population Sizes

We set the initial population sizes for each species to approximately half of the carrying capacity.

```{r initial-populations}
# Initial population sizes (approximately half of the carrying capacity)
initial_populations <- traits$Carrying_Capacity / 2

# Create a data frame to store population sizes over time
populations <- data.frame(
  Species = species,
  Year_0 = initial_populations
)

# Print initial populations
print(populations)
```

## Simulate Population Dynamics

We simulate population dynamics over 20 years, incorporating quantitative traits to calculate NPP and adding environmental stochasticity. We also incorporate the carrying capacity into the Lotka-Volterra model.

```{r simulate-population-dynamics}
# Number of years to simulate
years <- 20

# Simulate population dynamics and calculate NPP
for (year in 1:years) {
  new_populations <- sapply(1:10, function(i) {
    current_population <- populations[i, paste0("Year_", year - 1)]
    growth_rate <- traits$Growth_Rate[i]
    photosynthetic_efficiency <- traits$Photosynthetic_Efficiency[i]
    leaf_area_index <- traits$Leaf_Area_Index[i]
    water_use_efficiency <- traits$Water_Use_Efficiency[i]
    carrying_capacity <- traits$Carrying_Capacity[i]
    
    # Calculate NPP as a function of quantitative traits
    npp <- current_population * growth_rate * photosynthetic_efficiency * leaf_area_index * water_use_efficiency
    
    # Apply the population dynamics model with carrying capacity (Lotka-Volterra style)
    next_population <- current_population + npp * (1 - current_population / carrying_capacity)
    
    # Add environmental stochasticity
    next_population <- next_population + rnorm(1, 0, 0.2 * next_population)
    
    # Ensure population doesn't drop below 0
    next_population <- max(next_population, 0)
    
    return(next_population)
  })
  
  populations[, paste0("Year_", year)] <- new_populations
}

# Print simulated populations
print(populations)
```

## Prepare Data for Analysis

We reshape the data for analysis and merge it with both quantitative and qualitative traits.

```{r prepare-data}
# Reshape the populations data frame for easier analysis
populations_long <- populations %>%
  pivot_longer(cols = starts_with("Year_"), names_to = "Year", values_to = "Population") %>%
  mutate(Year = as.numeric(gsub("Year_", "", Year)))

# Merge traits with population data
data <- merge(populations_long, traits, by = "Species")

# Calculate NPP for each year
data <- data %>%
  mutate(NPP = Population * Growth_Rate * Photosynthetic_Efficiency * Leaf_Area_Index * Water_Use_Efficiency)

# Print prepared data
print(head(data))
```

## Fit Models

We fit three models: one without traits, one with qualitative traits, and one with quantitative traits.

### Model 1: Without Traits

```{r model-no-traits}
# Fit a linear mixed-effects model without traits
model_no_traits <- lmer(NPP ~ (1 | Species) + Year, data = data)
summary(model_no_traits)
```

### Model 2: With Qualitative Traits

```{r model-qualitative-traits}
# Convert qualitative traits to factors
data$Plant_Type <- as.factor(data$Plant_Type)
data$Shade_Tolerance <- as.factor(data$Shade_Tolerance)

# Fit a linear mixed-effects model with qualitative traits
model_qualitative_traits <- lmer(NPP ~ Plant_Type + Shade_Tolerance + (1 | Species) + Year, data = data)
summary(model_qualitative_traits)
```

### Model 3: With Quantitative Traits

```{r model-quantitative-traits}
# Fit a linear mixed-effects model with quantitative traits
model_quantitative_traits <- lmer(NPP ~ Growth_Rate + Photosynthetic_Efficiency + Leaf_Area_Index + Water_Use_Efficiency + (1 | Species) + Year, data = data)
summary(model_quantitative_traits)
```

## Model Comparison and Statistical Testing

We compare the models using AIC and conduct likelihood ratio tests.

```{r model-comparison}
# AIC values for the models
aic_no_traits <- AIC(model_no_traits)
aic_qualitative_traits <- AIC(model_qualitative_traits)
aic_quantitative_traits <- AIC(model_quantitative_traits)

# Print AIC values
print(paste("AIC without traits:", aic_no_traits))
print(paste("AIC with qualitative traits:", aic_qualitative_traits))
print(paste("AIC with quantitative traits:", aic_quantitative_traits))

# Conduct likelihood ratio tests
lr_test_qual_vs_no <- lrtest(model_qualitative_traits, model_no_traits)
lr_test_quant_vs_no <- lrtest(model_quantitative_traits, model_no_traits)
lr_test_quant_vs_qual <- lrtest(model_quantitative_traits, model_qualitative_traits)

# Print likelihood ratio tests
print(lr_test_qual_vs_no)
print(lr_test_quant_vs_no)
print(lr_test_quant_vs_qual)
```

## Visualizations

We visualize the simulated population dynamics and NPP for the species.

```{r plot-population-dynamics, fig.width=12, fig.height=8}
library(ggplot2)

# Plot population dynamics
ggplot(populations_long, aes(x = Year, y = Population, color = Species)) +
  geom_line() +
  theme_minimal() +
  labs(title = "Population Dynamics of 10 Species Over 20 Years", x = "Year", y = "Population Size")
```

```{r plot-npp, fig.width=12, fig.height=8}
# Plot NPP
ggplot(data, aes(x = Year, y = NPP, color = Species)) +
  geom_line() +
  theme_minimal() +
  labs(title = "NPP of 10 Species Over 20 Years", x = "Year", y = "Net Primary Productivity")
```

## Conclusion

By comparing models with no traits, qualitative traits, and quantitative traits, we demonstrate the utility of detailed trait data in explaining the variation in NPP. This analysis underscores the importance of functional traits in ecological research and management, providing insights into how different species contribute to ecosystem functions such as NPP.
