---
title: "plotting functional spaces"
format: html
embed-resources: true
toc: true
---

# 1. Set up

```{r}
#| message: false

library(tripack)
library(tidyverse)
library(here)

traits <- read_rds(here("data", "functional-traits", "trait_matrix.rds"))
comm_mat_algae <- read_rds(here("data", "SBC-LTE", "comm_mat_algae.rds"))
comm_meta <- read_rds(here("data", "SBC-LTE", "comm_meta.rds"))

low_col <- "#C70000"

medium_col <- "#54662C"

high_col <- "#114C54"
```

And just as a preview, this is what the trait matrix looks like:

```{r}
head(traits)
```

And the structure:

```{r}
str(traits)
```

And this is the community matrix:

```{r}
head(comm_mat_algae)
```

And this is the metadata:

```{r}
head(comm_meta)
```


# 2. Generate PCoA axes

Then, I generate the Gower dissimilarity matrix and do a PCoA to get the "synthetic" traits.

```{r}
# generate Gower matrix
trait_gower <- FD::gowdis(traits)

# doing PCoA to get dimensions
trait_pcoa <- ape::pcoa(D = trait_gower)

# extracting axis values as data frame
spp_pcoa_scores <- trait_pcoa$vectors %>% 
  as_tibble(rownames = NA) %>% 
  rownames_to_column("scientific_name")
```

# 3. Getting the "global pool"

So this is where I get the whole trait space for all the species in the study.

```{r}
# get the PCoA axes for axis
m <- trait_pcoa$vectors

# get the set of triangles that defines the hull
tr <- tri.mesh(m[,1], m[,2])

# get the points on the outside of the hull
# ch stands for "convex hull"
ch <- convex.hull(tr) %>% 
  data.frame() %>% 
  rownames_to_column("scientific_name")
```

And then just to see what the `ch` object looks like:

```{r}
ch
str(ch)
```


# 4. Survey trait space

So, I have a lot of surveys. I did this all in a nested data frame structure (instead of a for() loop) because I needed to do the same thing a bunch of times for different subsets of the data.  

The order of operations for each survey was:  
1. filter the species to only include the ones that actually occured during that survey  
2. get the PCoA axis values for those species  
3. take out the surveys that have fewer than 3 species  
4. calculate the convex hull for the surveys that have more than 3 species

At the end of this chunk, I'll have the convex hull for every survey (with more than 2 species, because you can't draw a convex hull with only two points, as I learned!).

```{r}
species_presence <- comm_mat_algae %>% 
  # turn the matrix into a data frame
  as_tibble(rownames = "sample_ID") %>% 
  
  # filter out species that don't appear in that survey
  # if the species biomass is greater than 0, replace that number with "yes"
  # if not, then "no"
  mutate(across(where(is.numeric), 
                ~ case_when(. > 0 ~ "yes", TRUE ~ "no"))) %>% 
  
  # make the data frame longer
  pivot_longer(cols = 2:43,
               names_to = "scientific_name",
               values_to = "presence") %>% 
  
  # join with PCoA scroes from above
  left_join(., spp_pcoa_scores, by = "scientific_name") %>% 
  
  # nest the data frame by sample_ID (representing each survey) 
  nest(.by = sample_ID, data = everything()) %>% 
  
  # NOTE: this is the start of the nested structure
  # filter the data frame to only include the species that occur in that survey
  mutate(filtered_comm = map2(
    data, sample_ID,
    ~ filter(.x, sample_ID == .y & presence == "yes")
  )) %>% 
  
  # create a matrix with the species name, PCoA 1 axis, and PCoA 2 axis
  mutate(coords = map(
    filtered_comm,
    ~ select(.x, 
             scientific_name, Axis.1, Axis.2) %>% 
      column_to_rownames("scientific_name") %>% 
      as.matrix()
  )) %>% 
  
  # count the number of species in that survey
  mutate(hull_outside_species = map(
    filtered_comm,
    ~ nrow(.x)
  )) %>% 
  
  # filter to only include surveys with more than 2 species
  filter(hull_outside_species > 2) %>% 
  mutate(trimesh = map(
    coords,
    ~ tri.mesh(.x[, 1], .x[, 2])
  )) %>% 
  
  # calculate the convex hull
  mutate(convex_hull = map(
    trimesh,
    ~ convex.hull(.x) %>% 
      data.frame() %>% 
      rownames_to_column("scientific_name")
  )) %>% 
  
  # join metadata information
  left_join(., comm_meta, by = "sample_ID")
```

# 5. wrangling

Whew! ok. So now I pull out the hulls for each survey and unnest the data frame.  

The output then looks like a regular data frame, but now with the species representing the outer edges of the convex hull (`scientific_name`) and the axis positions (`x` and `y`).

```{r}
hulls <- species_presence %>% 
  select(sample_ID, convex_hull) %>% 
  unnest(cols = c(convex_hull)) %>% 
  left_join(., comm_meta, by = "sample_ID")

head(hulls)
```

# 6. plotting

And then plot!

```{r}
during_space <- ggplot() +
  theme(panel.background = element_rect(fill = "lightgrey")) +
  
  # global species pool
  geom_polygon(data = ch,
               aes(x = x,
                   y = y),
               fill = "#FFFFFF") +
  geom_point(data = m,
             aes(x = Axis.1,
                 y = Axis.2),
             color = "grey",
             shape = 21,
             alpha = 0.5) +
  
  # survey hulls
  geom_polygon(data = hulls %>% 
                 filter(exp_dates == "during" & quality == "high"),
               aes(x = x,
                   y = y,
                   group = sample_ID),
               alpha = 0.025,
               fill = high_col
  ) +
  geom_polygon(data = hulls %>% 
                 filter(exp_dates == "during" & quality == "medium"),
               aes(x = x,
                   y = y,
                   group = sample_ID),
               alpha = 0.025,
               fill = medium_col
  ) +
  geom_polygon(data = hulls %>% 
                 filter(exp_dates == "during" & quality == "low"),
               aes(x = x,
                   y = y,
                   group = sample_ID),
               alpha = 0.03,
               fill = low_col
  ) +
  facet_grid(cols = vars(quality), vars(treatment)) +
  labs(title = "Removal period",
       x = "PCoA 1",
       y = "PCoA 2") +
  theme(strip.background = element_blank(),
        strip.text = element_text(size = 12),
        panel.grid = element_blank())

after_space <- ggplot() +
  theme(panel.background = element_rect(fill = "lightgrey")) +
  # global species pool
  geom_polygon(data = ch,
               aes(x = x,
                   y = y),
               fill = "#FFFFFF") +
  geom_point(data = m,
             aes(x = Axis.1,
                 y = Axis.2),
             color = "grey",
             shape = 21,
             alpha = 0.5) +
  geom_polygon(data = hulls %>% 
                 filter(exp_dates == "after" & quality == "high"),
               aes(x = x,
                   y = y,
                   group = sample_ID),
               alpha = 0.025,
               fill = high_col
  ) +
  geom_polygon(data = hulls %>% 
                 filter(exp_dates == "after" & quality == "medium"),
               aes(x = x,
                   y = y,
                   group = sample_ID),
               alpha = 0.025,
               fill = medium_col
  ) +
  geom_polygon(data = hulls %>% 
                 filter(exp_dates == "after" & quality == "low"),
               aes(x = x,
                   y = y,
                   group = sample_ID),
               alpha = 0.03,
               fill = low_col
  ) +
  facet_grid(cols = vars(quality), vars(treatment)) +
  labs(title = "Recovery period",
       x = "PCoA 1",
       y = "PCoA 2") +
  theme(strip.background = element_blank(),
        strip.text = element_text(size = 12),
        panel.grid = element_blank())

during_space
after_space
```

